---
title       : Early childhood care and education
subtitle    : Hartford regional data and indicators
author      : Scott Gaul
job         : Community Indicators Project
framework   : minimal        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
ext_widgets : {rCharts: ["libraries/nvd3"]} 
mode        : selfcontained # {standalone, draft}
---

## Early childhood care and education indicators

This review combines two indicators related to childcare, pre-K experience and school readiness that did well in the survey: 
* % 3-4 yr. olds in Head Start or other pre-school
* Subsidized childcare slots/1,000 children age 3-5

## Data and limitations

Data on early care and education is available from two primary sources. 

* The Census provides data on school enrollment through the American Community Survey (ACS). The ACS data is grouped by [age, race, type and level of school](http://factfinder2.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_11_5YR_B14003&prodType=table), but is only available for small geographies over a rolling 3 to 5 year period.
* The State Department of Education (SDE) provides data on pre-K experience and enrollment, along with data publicly-funded pre-K slots and data on kindergarten-readiness from the kindergarten inventory (KEI). 

One question is the extent to which these sources ask different questions and tell different stories about pre-K experience for this region. 

The Department of Social Services (DSS) also provides data on subsidies, such as Care4Kids, that are targeted at early childhood. 

## What do we already know?

Below are references for some local, regional and national reports that use data on early care and education. 

### Local

* [CT Kids Report Card](http://www.cga.ct.gov/kid/rba/results.asp) includes 'kindergarteners needing substantial help in school' as an indicator, using data from the SDE.
* [Health Equity Index](http://www.cadh.org/health-equity/health-equity-index.html) includes the percent of kindergarten students with pre-kindergarten educational experience.
* [Pioneer Valley State of the People](http://pvpc.org/resources/datastats/state-of-people/stateofthepeople2013.pdf) includes early education enrollment within their 'Education' domain. 
* [Data Haven's Community Well-Being Index](http://www.ctdatahaven.org/communityindex) includes the '% of children age 3-4 enrolled in preschool' as an indicator, using data from  the 2011 5-year American Community Survey estimates on school enrollment by age, along with data on the Kindergarten Inventory from SDE to assess school-readiness. 
* [Sustainable Knowledge Corridor Dashboard](http://www.sustainableknowledgecorridor.org/site/content/how-are-we-doing) includes 'Early education enrollment' from the American Community Survey.
* [Opportunities Hartford](http://www.cahs.org/programs-opportunitieshartford.asp) includes the number of children receiving Care4Kids subsidies under Early Care and Education. 
* [KidsCount](http://www.cahs.org/kidscount.asp) includes data on Care 4 Kids Child Enrollment from DSS and data on Pre-Kindergarten experience from SDE. 

### Other relevant efforts

* [The Child Well-Being Index](http://fcd-us.org/our-work/child-well-being-index-cwi) includes the rate of pre-kindergarten enrollment from national census data. 

## Sample results

### Enrollment in pre-K programs

The following results review data on pre-K experience and subsidized childcare and early childhood education slots from the various sources available. 

First, SDE provides data on the pre-K experience of students that enroll in kindergarten in the state. 

Aggregated data for the 30-town CRCOG region shows that roughly 75 percent of students enrolling in kindergarten have had some form of pre-K experience for the past few years. 

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9}
#Load libraries for graphs
library(reshape)
library(ggplot2)
library(scales)

#Load data
prek <- read.csv("preKSDE.csv")
#Filter out the districts that are not in the region and clear out the factors
prek <- subset(prek, prek$District.ID %in% levels(as.factor(districts$districtid)))

#Then add variable for Alliance Districts
prek$District.Group <- ifelse(prek$District.ID %in% c(11,43,47,64,77,146,164,165), "Alliance District","Not Alliance District")

prek_region <- ddply(prek,.(School.Year), summarise, value = weighted.mean(Percentage.of.Kindergarteners.with.PreK.Experience, Number.of.K.students))
ggplot(data = prek_region, aes(x = School.Year, y = value)) + 
  geom_line(aes(group = "")) +
  ylim(0,100) + 
  labs(x = NULL, y = NULL, title = "Pre-K experience rates, Greater Hartford") +
  theme_minimal()
```

Segmenting districts, the rates of pre-K experience are lower in the eight Alliance Districts, fluctuating between 50 to 75 percent, while higher-performing districts have had rates closer to 85 percent.

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9}
#Pre-K rates by town
prek_alliance <- ddply(prek,.(School.Year, District.Group), summarise, value = weighted.mean(Percentage.of.Kindergarteners.with.PreK.Experience, Number.of.K.students))

ggplot(data = prek_alliance, aes(x = School.Year, y = value)) + 
  geom_line(aes(group = District.Group, colour = District.Group)) +
  ylim(0,100) +
  labs(x = NULL, y = NULL) +
  theme_minimal()  
```

Most districts have fairly stable rates, although there are large year-on-year fluctuations in the SDE data for Hartford Public Schools and declining rates in some other districts. 

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height = 12}
#Pre-K rates by town
prek_dist <- ddply(prek,.(District.Name,School.Year), 
                   summarise,
                   value = weighted.mean(Percentage.of.Kindergarteners.with.PreK.Experience, Number.of.K.students))
prek_dist$District.Name <- reorder(prek_dist$District.Name, prek_dist$value)

ggplot(data = prek_dist, aes(x = School.Year, y = value)) + 
  geom_line(aes(group = District.Name)) +
  labs(x = NULL, y = NULL, title = "Pre-K experience rates, Greater Hartford") +
  scale_x_discrete(breaks = c("2002-2003","2006-2007","2010-11")) + 
  theme_minimal() +
  facet_wrap( ~ District.Name, ncol = 3)
```

For the last five years, Tolland Public Schools have the lowest rates of students with pre-K experience in the region, perhaps surprising given that it is among the higher-performing districts and more affluent towns in the region. 

```{r echo = FALSE, warning = FALSE, message = FALSE}
#Pre-K rates by town
prek_avg <- ddply(subset(prek_dist, School.Year %in% c("2010-11","2009-10","2008-09","2007-2008","2006-2007")), 
                  .(District.Name), summarise, value = mean(value))
prek_avg$District.Name <- reorder(prek_avg$District.Name, prek_avg$value)
ggplot(data = prek_avg, aes(x = value, y = District.Name)) + 
  geom_point() +
  labs(x = NULL, y = NULL, title = "Pre-K experience rates (2009-11), Greater Hartford") +
  theme_minimal()
```

While towns like Tolland are outliers, if we compare median household income and rates of pre-K experience, the general trend is still for higher rates of pre-K experience in higher income towns.

```{r echo = FALSE, warning = FALSE, message = FALSE}
#Create column with towns for district names
prek_avg$town= gsub(" School District", "", prek_avg$District.Name)
prek_avg <- merge(prek_avg, income, by = "town")

ggplot(data = prek_avg, aes(x = value, y = income)) + 
#  geom_point() +
  geom_text(aes(label = town)) +
  scale_y_continuous(labels = comma) + 
  geom_smooth() + 
  labs(x = "Pre-K experience rates", y = "Median household income") +
  theme_minimal()
```

The second primary data source on early education is the American Community Survey. The ACS includes data on school enrollment (public and private), [segmented by age, gender and other variables](http://factfinder2.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_12_1YR_B14003&prodType=table). 

The map below shows the percent of 3-4 year olds that are enrolled in public and private school programs, by town, for the five-year 2007 - 2011 period.

```{r echo = FALSE,warning = FALSE,message = FALSE, fig.width = 10, comment = NA, tidy = F}
library(ggplot2)
library(acs)
library(maps)
library(maptools)
key = "ba67d3a427e1f785987b9c8bc59341bf7c8a7cc1"
api.key.install(key)
#hartford.towns = geo.make(state = "CT", county = c("Hartford","Tolland"), county.subdivision = "*", check = T)
B14003 = acs.fetch(geography = hartford.towns, table.number = "B14003", col.names = "pretty")
prek.estimate = 
  divide.acs(numerator=(B14003[,4] + B14003[,13] + B14003[,32] + B14003[,41]),
             denominator=(B14003[,4] + B14003[,13] + B14003[,22] + B14003[,32] 
                          + B14003[,41] + B14003[,50]), method = 'proportion')
pop.estimate = B14003[,4] + B14003[,13] + B14003[,22] + B14003[,32] + B14003[,41] + B14003[,50]

prek.acs = data.frame(town=geography(B14003)[[1]],
                        prek=as.numeric(estimate(prek.estimate)),
                        se=standard.error(prek.estimate),
                      pop = as.numeric(estimate(pop.estimate))) 

names(prek.acs) <- c("town","prek","se","pop")
prek.acs$town= gsub(" town, Tolland County, Connecticut", "", prek.acs$town)
prek.acs$town= gsub(" town, Hartford County, Connecticut", "", prek.acs$town)
#Merge with data
library(classInt)
jenks <- classIntervals(prek.acs$prek, n=4, style="fisher")
choropleth=merge(CTTowns, prek.acs, by.x = "id", by.y="town")
choropleth=choropleth[order(choropleth$order), ]
choropleth$prek=cut(choropleth$prek, breaks=jenks$brks, include.lowest=T, dig.lab = T)
#Make the map
ggplot(data = choropleth, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = prek)) + 
  scale_x_continuous(breaks = NULL) +
  scale_y_continuous(breaks = NULL) +
  labs(x = NULL, y = NULL) + 
  coord_equal() +
  geom_polygon(data = CTTowns, colour = "grey", alpha = 0.5, fill = NA) +
  scale_fill_brewer(palette = "Purples", name = "3 - 4 year olds\nin school") +
  theme_minimal()
```

ACS results differ from the SDE data. Tolland (for instance) has one of the highest rates in the ACS data, and Hartford is closer to the middle of the pack. According to the ACS data, urban periphery towns like East Hartford, Enfield and Bloomfield have the lowest rates of enrollment in the region.

Since the data is drawn from the ACS, there are also margins of error associated with the values, which can be large for small towns.  

```{r echo = FALSE, warning = FALSE, message = FALSE}
prek.acs <- merge(prek_avg, prek.acs, by = "town")
prek.acs$town <- reorder(prek.acs$town, prek.acs$prek)
library(scales)
ggplot(data = prek.acs) + 
  geom_point(aes(x = prek, y = town)) + 
  geom_segment(aes(xend = prek - se.y, x = prek + se.y, y = town, yend = town), 
               colour = "grey") + 
  labs(x = '% of 3 - 4 year olds enrolled in school', y = NULL) + 
  scale_x_continuous(labels = percent) + 
  theme_minimal()
```

Direct comparison shows where 5-year average rates from ACS and SDE agree and disagree.  Each measures slightly different things - the ACS data measures the number of 3 - 4 year olds living in these towns that attend school. The SDE data measures the number of children attending kindergarten in the respective school districts that have some prior pre-kindergarten experience. 

Which data source does a better job identifying the important issues for our region? 

```{r echo = FALSE, warning = FALSE, message = FALSE}
ggplot(data = prek.acs, aes(x = value / 100, y = prek)) + 
#  geom_point() +
  geom_text(aes(label = town)) +
  scale_y_continuous(labels = percent, limits = c(0.5,1)) + 
  scale_x_continuous(labels = percent, limits = c(0.5,1)) + 
  geom_smooth(method = lm, se = FALSE) + 
  labs(x = "Pre-K experience rates (SDE)", y = "Pre-K experience rates (ACS)") +
  theme_minimal()
```

### Subsidies for pre-K programs

The next set of indicators are on subsidies for pre-K and other childcare programs. 

SDE publishes data on publicly-funded pre-K slots. Not all towns have available slots, but for those with slots, Bloomfield has the highest rate relative to the size of the 3 - 4 year-old population in the region. 

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9}
#Load data on publicly-funded preK-slots
publicslots <- read.csv("publicslots.csv")
publicslots <- subset(publicslots, publicslots$Location %in% levels(towns$Town))
publicslots$Location <- factor(publicslots$Location)
publicslots <- ddply(publicslots, .(Location), summarise, 
                     Number.of.Students = mean(Number.of.Students))
publicslots <- merge(publicslots, prek.acs, by.x = "Location", by.y = "town")
publicslots$Slots.per.Kid = publicslots$Number.of.Students / publicslots$pop
publicslots$Location <- reorder(publicslots$Location, publicslots$Slots.per.Kid)
#Group and merge by town ID and district ID
ggplot(data = publicslots) + 
  geom_point(aes(x = Slots.per.Kid, y = Location)) + 
  scale_x_continuous(labels = percent) + 
  labs(y = NULL, x = "Average # of publicly funded slots as percent of 3 - 4 year olds population, Greater Hartford") +
  theme_minimal()
```

The publicly-funded childcare slots are also available by funding source and with several demographic breakouts. 

Care 4 Kids data is also available in [the most recent KidsCount report](http://www.cahs.org/pdf/CAHS2013-KIDSCOUNT-ALLPGS-SINGLES.pdf). The report notes that there are 'no surprises' - urban centers have higher enrollment, although some inner-ring suburbs also have high enrollment rates. 

In this case, Hartford has the highest rate in the region, followed by Bloomfield and East Hartford where the average number enrolled is more than half of all 3 - 4 year olds living in the town. 

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9}
#Load data on Care 4 Kids enrollment
care4kids <- read.csv("care4kids.csv")
care4kids <- ddply(care4kids, .(Town), summarise, Enrollment = mean(Enrollment))
care4kids <- merge(care4kids, prek.acs, by.x = "Town", by.y = "town")
care4kids$Enrollment.Rate = care4kids$Enrollment / care4kids$pop
care4kids$Town <- reorder(care4kids$Town, care4kids$Enrollment.Rate)
ggplot(data = care4kids) + 
  geom_point(aes(x = Enrollment.Rate, y = Town)) + 
  scale_x_continuous(labels = percent) + 
  labs(y = NULL, x = "Average Care 4 Kids enrollment (2005 - 2012) as percent of 3-4 year-old population, Greater Hartford") +
  theme_minimal()
```


